---
title: "Assignment 2"
format: html
editor: visual
---

## Loading in the required packages and data to be used to answer the problems.

```{r}
# Loading required packages.
library(tidyverse)  # loads ggplot2, dplyr, readr
library(lubridate)  # for date/time manipulation

```

```{r}
# Loading in the data
df <- read_csv("elmarket_13_19.csv")
```

## Assignment 2

## Task A. Plotting monthly averages of temparature, volume, price and wind production.
```{r}
# Creating a function that plots the average monthly data of a specified variable.
plot_avg_monthly <- function(data, variable) {
  
  # Check if the specified variable is valid
  if (variable %in% names(data)) {
    
    # Extract year and month, calculate the average per month
    avg_data <- data %>%
      mutate(year_month = floor_date(date, "month")) %>%
      group_by(year_month) %>%
      summarise(avg_var = mean(get(variable), na.rm = TRUE))
    
    # Plot
    p <- ggplot(avg_data, aes(x = year_month, y = avg_var)) +
      geom_line() +
      labs(y = paste("Average", variable),
           x = "Month",
           title = paste("Average Monthly", variable)) +
      theme_classic()
    return(p)
    
  # If the specified variable is invalid  
  } else {
    stop("Invalid variable name. Choose a variable present in the data frame.")
  }
}
```

```{r}
# Create a loop that creates plots of specified variables

# List of variables that we want to visualize monthly data
variable_list <- c("temperature", "volume", "price", "wind_production")

# Initialized list to store the plot objects
plot_list <- list()

# Looping through the variable list and applying the plot function for each variable
for (i in variable_list) {
  p <- plot_avg_monthly(data = df, variable = i)  # calling the monthly plot function
  plot_list[[i]] <- p                             # storing the plot in a list making it accessible later                 
}
```

We start by commenting the plot of monthly averages of temperature.
```{r}
# Plot monthly average temperature
plot_list$temperature
```
As expected we see variations in average temperatures between the seasons. We notice that the lowest temperatures are recorded during winter and the highest during summers. Also, we notice that the min. and max. temperatures are relatively equal each year.

We can also comment on the plot of monthly averages in volume. 
```{r}
# Plot monthly average volume
plot_list$volume
```
In the plot we se that the energy consumption is is at max during winter and are lowest during winters. There are also differences between different year, but no great differences.

The next plot to examine is the plot of monthly averages of price.
```{r}
# Plot monthly average price
plot_list$price
```
We notice some seasonal trends in the prices of energy, however, the biggest variations is betwween years. W notice a substantial dip in energy prices during the summer of 2015, however the price continued to rise up until the winter of 2019. After that it has returned to 2018-levels.

The last single plot to examine is the plot of monthly wind production.
```{r}
# Plot monthly average wind production
plot_list$wind_production. 
```
Also in this plot we notice seasonal variations with highs during autumn/winter and lows during summer.


```{r}
# Create a function that handles multiple variables and creates single plot
plot_avg_monthly_multi <- function(data, variables) {
  
  # Check if the specified variable is valid
  if (all(variables %in% names(data))) {
    
    # Extract year and month, calculate the average per month of both variables
    avg_data <- data %>%
      mutate(year_month = floor_date(date, "month")) %>%
      group_by(year_month) %>%
      summarise(avg_var1 = mean(get(variables[1]), na.rm = TRUE),
                avg_var2 = mean(get(variables[2]), na.rm = TRUE), 
                .groups = "drop")
    
    # Compute a scaling factor to map both variables onto a similar scale
    scale_factor <- mean(avg_data$avg_var1, na.rm = TRUE) / 
                    mean(avg_data$avg_var2, na.rm = TRUE)
    
    # Plot
    p <- ggplot(avg_data, aes(x = year_month)) +
      geom_line(aes(y = avg_var1), color = "red") +
      geom_line(aes(y = avg_var2 * scale_factor), color = "blue", linetype = "dashed") +
      scale_y_continuous(
        name = paste("Average", variables[1]),
        sec.axis = sec_axis(~. / scale_factor, name = paste("Average", variables[2])),
        labels = scales::number
      ) +
      labs(
        x = "Month",
        title = paste("Monthly averages of", paste(variables, collapse = " & "))
      ) +
      theme_classic()
    return(p)
    
  # If the specified variable is invalid  
  } else {
    stop("Invalid variable name. Choose variables present in the data frame.")
  }
}
```

```{r}
# Plot relationship between variables

# Initialize empty list
plot_list_multi <- list()

# Loop over variables and store comparison plots in list
for (i in 1:(length(variable_list) - 1)) {
  for (j in (i + 1):length(variable_list)) {
    current_vars <- c(variable_list[i], variable_list[j])
    p <- plot_avg_monthly_multi(df, current_vars)
    plot_name <- paste(current_vars, collapse = "_vs_")
    plot_list_multi[[plot_name]] <- p
  }
}
```

