---
title: "Assignment 2"
format: html
editor: visual
---

## Loading in the required packages and data to be used to answer the problems.

```{r}
# Loading required packages.
library(tidyverse)  # loads ggplot2, dplyr, readr
library(lubridate)  # for date/time manipulation

```

```{r}
# Loading in the data
df <- read_csv("elmarket_13_19.csv")
```

## Assignment 2

## Task A. Plotting monthly averages of temparature, volume, price and wind production.
```{r}
# Creating a function that plots the average monthly data of a specified variable.
plot_avg_monthly <- function(data, variable) {
  
  # Check if the specified variable is valid
  if (variable %in% names(data)) {
    
    # Extract year and month, calculate the average per month
    avg_data <- data %>%
      mutate(year_month = floor_date(date, "month")) %>%
      group_by(year_month) %>%
      summarise(avg_var = mean(get(variable), na.rm = TRUE))
    
    # Plot
    p <- ggplot(avg_data, aes(x = year_month, y = avg_var)) +
      geom_line() +
      labs(y = paste("Average", variable),
           x = "Month",
           title = paste("Average Monthly", variable)) +
      theme_classic()
    return(p)
    
  # If the specified variable is invalid  
  } else {
    stop("Invalid variable name. Choose a variable present in the data frame.")
  }
}

# List of variables that we want to visualize monthly data
variable_list <- c("volume", "temperature", "price", "wind_production")

# Initialized list to store the plot objects
plot_list <- list()

# Looping through the variable list and applying the plot function for each variable
for (i in variable_list) {
  p <- plot_avg_monthly(data = df, variable = i)  # calling the monthly plot function
  plot_list[[i]] <- p                             # storing the plot in a list making it accessible later                 
}
```

We start by commenting the plot of monthly averages of temperature.
```{r}
# Plot monthly average temperature
plot_list$temperature
```
As expected we see variations in average temperatures between the seasons. We notice that the lowest temperatures are recorded during winter and the highest during summers. Also, we notice that the min. and max. temperatures are relatively equal each year.

We can also comment on the plot of monthly averages in volume. 
```{r}
# Plot monthly average volume
plot_list$volume
```

